Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-11-16 paper/paper.tex
a few tiny tweaks

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

@@

\subsection{A system to identify and enrich for rare IFN+ cells}
A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
We previously performed single-cell transcriptomics on influenza-infected A549 human lung epithelial cells, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN~\citep{russell2018extreme}.
Prior studies using reporters have also found that IFN expression is rare in influenza-infected cells~\citep{killip2017single} and mice~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
Reporter cells to identify and enrich IFN+ cells.
{\bf (A)} An A549 IFN reporter cell line.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \SUPPFIG{reportervalidation}C for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

@@

%%% start workflow figure
\begin{figure}
\begin{fullwidth}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.

>>>>

\section{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

@@

\subsection{A system to identify and enrich for rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}[b!]
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
Reporter cells to identify and enrich IFN+ cells.
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (\SUPPFIG{reportervalidation}C).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).

@@

%%% start workflow figure
\begin{figure}[b!]
\begin{fullwidth}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
